A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-29 paper/writeup.tex
Tweak

First proposed by~\cite{kuhner2015assessing}, a common approach 
is to use a weighted average Robinson-Foulds (RF) distance
% Other authors \citep[e.g.,][]{kelleher2019inferring} have used
% an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between ARGs.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
the indices of all trees on which there is \emph{some} node in $\T_2$
whose set of descendant samples matches those of $n$, and
\begin{align*}
    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)}^N (a_k - a_{k-1})
\end{align*}
the total span over which $n$ finds a match.
Then the average RF distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average RF distance allows a different node to match
on each tree.
The other differences are that \emph{average} RF distance
normalizes by sequence length, and is symmetrized.
The RF distance between two trees
was defined by \citet{robinson1981comparison}
% JK - change edge->branch here to avoid confusion with tskit/ARG edge
to be the minimum number of branch contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges 
that induce different splits on the labels).
A similar metric on ARGs could be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

>>>>

A common measure of disagreement between ARGs, 
first proposed by~\cite{kuhner2015assessing}, 
is to use a weighted average Robinson-Foulds (RF) distance.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
the indices of all trees on which there is \emph{some} node in $\T_2$
whose set of descendant samples matches those of $n$, and
\begin{align*}
    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)}^N (a_k - a_{k-1})
\end{align*}
the total span over which $n$ finds a match.
Then the average RF distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average RF distance allows a different node to match
on each tree.
The other differences are that \emph{average} RF distance
normalizes by sequence length, and is symmetrized.
The RF distance between two trees
was defined by \citet{robinson1981comparison}
% JK - change edge->branch here to avoid confusion with tskit/ARG edge
to be the minimum number of branch contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges 
that induce different splits on the labels).
A similar metric on ARGs could be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.
